Trail Saving on Backtrack

### 2020-02-27 paper.tex
typos

In this paper we present an alternative ``trail savings'' method whereby
we save the backtracked part of the solver's trail and attempt to
use that information to make the solver's redescent more
efficient. Trail saving preserves the traditional invariants of the
SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and sometimes better than
chronological backtracking. We also show that with the enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
which is equal to the number of decision literals appearing before
$\ell$ on the trail plus one; hence, $\dlevel{d}=1$ for the first
decision literal $d\in\trail$. The set of literals on $\trail$ that
have the same decision level forms a \emph{contiguous}
subsequence\footnote{Our approach uses standard trails in which the
  decision levels are contiguous. Chronological backtracking
  \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
  generates trails with non-contiguous decision levels} that starts
with a decision literal $d_i$ and ends just before the next decision
literal $d_{i+1}$. We will often need to refer to different decision
level subsequences of $\trail$. Hence, we let $\trail[[i]]$ denote the
subsequence of literals at decision level $i$; and let
$\trail[[\range{i}{j}]]$ denote the subsequence of literals at
decision levels $k$ for $i\leq k\leq j$. 
\begin{defn}
    A clause $C$ has \textbf{been made unit by $\trail$ implying $l$}
    when
    $l\in C \land \bigl(\forall x\in C. x\neq l \implies \lnot x \in
    \trail\bigr)$. That is, all literals in $C$ except $l$ must have
    been falsified by $\trail$
\end{defn}

Now we define the following properties that a trail $\trail$ can have.
\begin{description}
\item[non-contradictory:] A variable cannot appear in both polarities
    in the trail: $l\in \trail\implies \lnot l \not \in \trail$.
\item[non-redundant:] A literal can only appear once on $\trail$.
\item[reason-sound:] For each implied literal $l\in\trail$ we have
    that its reason clause $\reason{l}=C$ has been made unit by
    $\trail$ implying $l$, and for each $x\in C$ with $x\neq l$ we
    that $\lnot x$ appears before $l$ on $\trail$:
    $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
    l\in\reason{l} \land \bigl(\forall x\in\reason{l}.\, x\neq l
    \implies \lnot x \in \trail \land \trailIdx{\lnot x} <
    \trailIdx{l}\bigr)$.
\item[propagation-complete:] Unit propagation has been run to
    completion at all decisions levels of $\trail$. This means that
    literals appear on $\trail$ at the first decision level they were
    unit implied. Formally, this can be captured by the condition:
    $\forall i \in \{\dlevel{l}\,|\,l\in \trail\}. \bigl(\exists
    C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. Note
    that propagation completeness implies that
    $\reason{l}\neq \varnothing$ must contain at least one other
    literal $y\neq l$ with $\dlevel{y} = \dlevel{l}$.
\item[conflict free:] No clause of $F$ is falsified by
    $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
    called \emph{conflicts}.
\end{description}

@@

\begin{example}
    For example, say that $(x, \lnot y)\in \formula$, the literal $x$
    is a decision literal on the trail with $\dlevel{x}=2$, and that
    the solver is currently at level $150$ where it encounters a
    conflict. If this conflict yields the unit clause $(y)$, standard
    backtracking would backtrack to level $0$, where $x$ would be
    implied. On redescent, $x$ would no longer form a new decision
    level and it would not appear in any new clauses (as it is
    entailed by $\formula$). \cbt, on the other hand, would backtrack
    to level $149$. On its trail $x$ would still be at level
    $2$. Until a backtrack past level $2$ occurs, learnt clauses might
    contain $\lnot x$, and thus have level $2$ added to their set of
    levels (potentially changing their LBD score). Only when backtrack
    past level $2$ occurs would $x$ be restored to its correct level
    $0$, and it would require inprocessing simplifications to remove $x$
    from the learnt clauses.
\end{example}

In sum, although these impacts of \cbt might or might not be harmful
to the \sat solver, they exist. Hence, perhaps the most unappealing
aspect of \cbt is that introduces additional inter-dependencies
between components of the solver. For example, the second impact means
that heuristics for choosing decision variables might work differently
in the presence of \cbt; and the first impact means that different
ways of computing clause quality using the decision levels in the
clauses might also work differently in the presence of \cbt. So
although there are bound to be cases where \cbt is superior, we would
like to have a method that addresses the original motivation of
reducing redundant work while not changing the operation of standard
backtracking. We present our proposal for such a method next.

@@

Say that the solver is at $\deepestLvl$ where it has encountered a
conflict. From the 1-UIP clause it learns, $\uipcls$, it now has to
backtrack to $\btL$. This is accomplished by calling
\textsc{backtrack($\btL$)}, shown in Figure~\ref{fig:trailUse}, which
saves the backtracked portion of the trail.

@@

After backtrack the solver will add $\deepestLit$ to the end of the
updated $\trail$ with $\reason{\deepestLit}$ $\mbox{}=\uipcls$ and
then invoke unit propagation. $\trailsave$ is exploited during
propagation by the version of \textsc{propagate} shown in
Figure~\ref{fig:trailUse}, which will initially be invoked with the
argument $\trailIdx{\deepestLit}$ (i.e., the trail index of the newly
added implicant). The saved trail be continually consulted during the
solver's descent whenever unit propagation is performed. When
backtrack occurs $\trailsave$ will be overwritten to store the new
backtracked portion of $\trail$.

@@

\begin{figure}[!t]
\begin{tabular}{l
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}}
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $l_5^{3*}$ & $l_6^3$ & $l_7^{4*}$ & $l_8^4$
      & $l_9^{5*}$ & $l_{10}^5$ & $l_{11}^5$ & $l_{12}^{6*}$
      & $l_{13}^6$ & $l_{14}^6$  \tabularnewline
     $\trailsave$
      &  &  &  & 
      &  &  &  &  
      &  &  &  &  
      &  & \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{1-UIP ($\lnot l_1$, $\lnot l_3$, $\lnot l_{12}$) and backtrack 2 ($\btL$)}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      &         &       &         &      
      &         &       &         &           
      &         &         \tabularnewline
      $\trailsave$
      &         &       &         &                     
      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
      & $l_9^*$ & $l_{10}$ & $l_{11}$ & 
      &         & 
      \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{Unit Prop $\lnot l_{12}$ ($\deepestLit$), $\trailsave$ not yet helpful}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $\smash{\lnot} l_{12}^2$  & $l_7^2$  & $l_9^2$    &      
      &         &       &         &           
      &         &         \tabularnewline
      $\trailsave$
      &         &       &         &                     
      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
      & $l_9^*$ & $l_{10}$ & $l_{11}$ &
      &         & 
      \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{Make decision}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $\smash{\lnot} l_{12}^2$  & $l_7^2$  & $l_9^2$ 
      & $l_5^{3*}$  &       &         &           
      &         &         \tabularnewline
      $\trailsave$
      &         &       &         &                     
      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
      & $l_9^*$ & $l_{10}$ & $l_{11}$ & 
      &         & 
      \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{Now $\trailsave$ can be used to augment trail}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $\lnot l_{12}^2$ \hfill & $l_7^2$  & $l_9^2$      
      & $l_5^{3*}$  & $l_6^{3}$  & $l_8^3$  & $l_{10}^3$          
      & $l_{11}^3$  &         \tabularnewline
      $\trailsave$
      &         &       &         &      
      &         &       &                              
      &         &       &         & 
      &         &       & 
    \end{tabular}
    \caption{Use of $\trailsave$ from
      Example~\ref{ex:trailsave}\label{fig:trailsave}. The literals
      decision level is indicated in its superscript, and a $\mbox{}^*$
      superscript indicates that the literal is a decision.}
\end{figure}

\begin{example}
    \label{ex:trailsave} Figure~\ref{fig:trailsave} provides an
    example of how $\trailsave$ is used. Initially the literals $l_1$
    to $l_{14}$ are on the solvers $\trail$, and $\trailsave$ is
    empty. This is shown in the first two lines of the figure. In the
    figure the superscript on the literals indicates their decision
    level, and a superscripted $*$ indicates that the literal is a
    decision. Hence $l_1^{1*}$ indicates that $\dlevel{l_1}=1$ and
    that $l_1$ is a decision.

    Then a conflict is found at level $6$ and the 1-UIP clause
    $(\lnot l_1, \lnot l_3, \lnot l_{12})$ is learnt. Thus the solver
    will backtrack to level $2$, where it will add $\lnot l_{12}$ as a
    unit implicant. The next two lines shows $\trail$ and $\trailsave$
    right after the backtrack to level $2$: the backtracked levels
    have been copied into $\trailsave$ omitting the conflict level $6$.

@@

As noted in the previous example unit propagation has to be rerun on
all saved literals added to $\trail$ from $\trailsave$. Thus our
technique, unlike \cbt, does not completely remove the overhead of
reproducing the trail on the solver's redescent. Nevertheless, trail
savings improves the efficiency of this redescent in three different
ways. First, by adding more forced literals to the trail before
continuing propagating the next literal, propagation can potentially
gain a quadratic speedup
\cite{DBLP:conf/sat/HickeyB19,DBLP:journals/jair/Gent13}. Second,
propagation does not need to examine the reason clause of the added
literals. If these literals were not added by \ust, propagation would
have to traverse each of these reason clauses to determine that they
are have in fact become unit. Third, when a conflict is returned by
\ust all further propagations can be halted. The added literals and
their reasons will be sufficient to perform clause learning from the
conflict returned by \ust. Since trail saving can sometimes save
hundreds or thousands of literals at a time these savings can in sum
be significant.

@@

\subsection{Correctness}
Now we will prove that our use of $\trailsave$ preserves the \sat
solver's soundness. In particular, $\trailsave$ is only used in the
procedure \ust, in which it either adds new literals
to the solver's trail, or returns conflict clauses to the solver.
Hence, we only need to show that these new literals are in fact unit
implied and the conflicts are in fact falsified by the solver's trail.
\
Since both $\trail$ and $\trailsave$ are sequences of literals (with
associated reasons) we can consider their concatenation denoted as
$\trail + \trailsave$.
\begin{theorem}
    \label{thm:sound}
    If $\trail + \trailsave$ is \textbf{reason sound}
    (Section~\ref{sec:background}) then the following holds.  If the
    first $i$ literals on $\trailsave$ are all in $\trail$
    ($\forall j. 0\leq j< i. \trailsave[i]\in \trail$) and
    $\trailsave[i]=l$ is an implied literal with $\reasonsave{l} = C$,
    then $C$ has been made unit by $\trail$ implying $l$.
\end{theorem}
\noindent\emph{Proof:}
Since $\trail + \trailsave$ is reason sound, every literal in $C$
other than $l$ appears negated before $l$ in the sequence
$\trail + \trailsave$. Thus for $x\in C$ we have $\lnot x \in \trail$
or $\lnot x \in \trailsave[0]\ldots\trailsave[i-1]$. But in the later
case we also $\lnot x \in \trail$. 
\whitebox

@@

\begin{theorem}
    \label{thm:rsound}
    $\trail+\trailsave$ is always reason sound during the operation of
    the solver.
\end{theorem}
\noindent \emph{Proof:} $\trailsave$ starts off being empty, so
$\trail+\trailsave = \trail$ is reason sound as it was generated by
the solver (4). In procedure \bt $\trail+\trailsave$ is
set to a trail that was previously generated by solver (4). The solver
can add to $\trail$ by decisions and propagations with out using
$\trailsave$. In this case $\trail' = \trail+\trail_{\mathit{new}}$,
and $\trail'$ is reason sound by (4), thus the new
$\trail'+\trailsave$ is reason sound by (3). Finally, in procedure
\ust either (a) literals at the front of
$\trailsave$ are discarded since they already appear on $\trail$, or
(b) literals are moved from $\trailsave$ to $\trail$. Under both of
these changes $\trail+\trailsave$ remains reason sound by (1) and (2).
\whitebox

@@

\paragraph{Saving the trail over multiple backtracks.}
It can often be the case that when the solver finds a conflict and
backtracks to $\btL$ it might immediately find a another conflict at
$\btL$ causing a further backtrack. In the procedure \bt every
backtrack causes $\trailsave$ to be overwritten. Hence, in these cases
most of the trail will not be saved---only the portion from the last
backtrack. Our first extension addresses this potential issue and also
provides more general trail savings in other contexts as well.

@@

\paragraph{Lookahead for Conflicts.}
In \ust we stop adding literals from $\trailsave$ to $\trail$ once we
reach a decision literal $d$ on $\trailsave$ that is not yet on
$\trail$ (line~\ref{ln:decnottrue} of \ust). This is done so that the
solver has full control over variable decisions without interference
from the trail saving mechanism (unlike the case with \cbt). However,
another option would be to force the solver to use $d$ as its next
decision literal, which would then allow us to further add all of
$d$'s implied literals on $\trailsave$ onto $\trail$. This can be done
for the first $k$ decisions on $\trailsave$ for any $k$. But in
general, we do not want to remove the solver's autonomy by forcing it
to made potentially different decisions than it might have wanted to.

@@

We experimented with this ``lookahead for conflicts'' idea using
various values of $k$. We found that $k=2$, i.e., forcing up to two
decisions from $\trailsave$ to be made by the solver if this yields a
conflict, often enhanced the solver's performance. Limiting the lookahead to
only one decision level of $\trailsave$ was not as good, and looking
ahead more than 2 decisions of $\trailsave$ also degraded
performance. This is provides some evidence that taking too much
control away from the solver and forcing it to make too many decisions
from $\trailsave$ can lead to conflicts that are not as useful to the
solver.

\paragraph{Reason quality.}
The saved trail can be thought of as remembering the solver's recent
trajectory. Sometimes we want to follow the past trajectory, but
perhaps sometimes we don't. In particular, when adding literals from
$\trailsave{}$ to $\trail$ we can examine the quality of the saved
reasons to see if they are worth using. Once we encounter a literal
with a low quality saved reason we stop adding literals from
$\trailsave$ to the solver's trail. In particular, we can change
lines~\ref{ln:addToTrail} and \ref{ln:addToTrailNext} of \ust
to the following:\\
\begin{tabular}[t]{l}
23.5 \qquad \textbf{if} (lowQuality($\reasonsave{\litsave}$)) \textbf{break}\\
24.  \qquad $\trail.\mathrm{addToEnd(\litsave)}$\\
25.  \qquad $\reason{\litsave}\gets \reasonsave{\litsave}$
\end{tabular}

Note that the solver will still set the un-added literals as they are
unit implied by $\trail$, but it might be able to find better reasons
for these implicants. There is of course no guarantee that better
reasons will be found, but our empirical results show that sometimes
this does happen. We experimented with two quality metrics, clause
size and clause LBD, obtaining positive results with both. These
results also provides evidence against the argument given in
\cite{DBLP:journals/jsat/TakRH11} that changing literal reasons is not
impactful. With an appropriate clause quality metric the changing literal
reasons can have an impact.

@@

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      maple           & 458          & 259 &  199  & 4746                             \\ \hline
      maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
      maple-trail-multipleBT-lookahead-reason & 999 & 999 & 999 & 9999 \\ \hline
    \end{tabular}
    \caption{Table of results for MapleLCMDist.}
    \label{fig:maple}
\end{figure}

>>>>

In this paper we present an alternative ``trail saving'' method whereby
we save the backtracked part of the solver's trail and attempt to
use that information to make the solver's redescent more
efficient. Trail saving preserves the traditional invariants of the
SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and sometimes better than
chronological backtracking. We also show that with the enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
which is equal to the number of decision literals appearing on the trail
up to and including $\ell$; hence, $\dlevel{d}=1$ for the first
decision literal $d\in\trail$. The set of literals on $\trail$ that
have the same decision level forms a \emph{contiguous}
subsequence\footnote{Our approach uses standard trails in which the
  decision levels are contiguous. Chronological backtracking
  \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
  generates trails with non-contiguous decision levels} that starts
with a decision literal $d_i$ and ends just before the next decision
literal $d_{i+1}$. We will often need to refer to different decision
level subsequences of $\trail$. Hence, we let $\trail[[i]]$ denote the
subsequence of literals at decision level $i$; and let
$\trail[[\range{i}{j}]]$ denote the subsequence of literals at
decision levels $k$ for $i\leq k\leq j$. 
\begin{defn}
    A clause $C$ has \textbf{been made unit by $\trail$ implying $l$}
    when
    $l\in C \land \bigl(\forall x\in C. x\neq l \implies \lnot x \in
    \trail\bigr)$. That is, all literals in $C$ except $l$ must have
    been falsified by $\trail$
\end{defn}

Now we define the following properties that a trail $\trail$ can have.
\begin{description}
\item[non-contradictory:] A variable cannot appear in both polarities
    in the trail: $l\in \trail\implies \lnot l \not \in \trail$.
\item[non-redundant:] A literal can only appear once on $\trail$.
\item[reason-sound:] For each implied literal $l\in\trail$ we have
    that its reason clause $\reason{l}=C$ has been made unit by
    $\trail$ implying $l$, and for each $x\in C$ with $x\neq l$ we
    have that $\lnot x$ appears before $l$ on $\trail$:
    $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
    l\in\reason{l} \land \bigl(\forall x\in\reason{l}.\, x\neq l
    \implies \lnot x \in \trail \land \trailIdx{\lnot x} <
    \trailIdx{l}\bigr)$.
\item[propagation-complete:] Unit propagation has been run to
    completion at all decisions levels of $\trail$. This means that
    literals appear on $\trail$ at the first decision level they were
    unit implied. Formally, this can be captured by the condition:
    $\forall i \in \{\dlevel{l}\,|\,l\in \trail\}. \bigl(\exists
    C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. Note
    that propagation completeness implies that
    $\reason{l}\neq \varnothing$ must contain at least one other
    literal $y\neq l$ with $\dlevel{y} = \dlevel{l}$.
\item[conflict-free:] No clause of $F$ is falsified by
    $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
    called \emph{conflicts}.
\end{description}

@@

\begin{example}
    For example, suppose that $(x, \lnot y)\in \formula$, the literal $x$
    is a decision literal on the trail with $\dlevel{x}=2$, and that
    the solver is currently at level $150$ where it encounters a
    conflict. If this conflict yields the unit clause $(y)$, standard
    backtracking would backtrack to level $0$, where $x$ would be
    implied. On redescent, $x$ would no longer form a new decision
    level and it would not appear in any new clauses (as it is
    entailed by $\formula$). \cbt, on the other hand, would backtrack
    to level $149$. On its trail $x$ would still be at level
    $2$. Until a backtrack past level $2$ occurs, learnt clauses might
    contain $\lnot x$, and thus have level $2$ added to their set of
    levels (potentially changing their LBD score). Only when backtrack
    past level $2$ occurs would $x$ be restored to its correct level
    $0$, and it would require inprocessing simplifications to remove $x$
    from the learnt clauses.
\end{example}

In sum, although these impacts of \cbt might or might not be harmful
to the \sat solver, they exist. Hence, perhaps the most unappealing
aspect of \cbt is that it introduces additional interdependencies
between components of the solver. For example, the second impact means
that heuristics for choosing decision variables might work differently
in the presence of \cbt; and the first impact means that different
ways of computing clause quality using the decision levels in the
clauses might also work differently in the presence of \cbt. So
although there are bound to be cases where \cbt is superior, we would
like to have a method that addresses the original motivation of
reducing redundant work while not changing the operation of standard
backtracking. We present our proposal for such a method next.

@@

Suppose that the solver is at $\deepestLvl$ where it has encountered a
conflict. From the 1-UIP clause it learns, $\uipcls$, it now has to
backtrack to $\btL$. This is accomplished by calling
\textsc{backtrack($\btL$)}, shown in Figure~\ref{fig:trailUse}, which
saves the backtracked portion of the trail.

@@

After backtrack the solver will add $\deepestLit$ to the end of the
updated $\trail$ with $\reason{\deepestLit}$ $\mbox{}=\uipcls$ and
then invoke unit propagation. $\trailsave$ is exploited during
propagation by the version of \textsc{propagate} shown in
Figure~\ref{fig:trailUse}, which will initially be invoked with the
argument $\trailIdx{\deepestLit}$ (i.e., the trail index of the newly
added implicant). The saved trail will be continually consulted during the
solver's descent whenever unit propagation is performed. When
backtrack occurs $\trailsave$ will be overwritten to store the new
backtracked portion of $\trail$.

@@

\begin{figure}[!t]
\begin{tabular}{l
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}
  >{\raggedleft}p{4.75ex}>{\raggedleft}p{4.75ex}}
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $l_5^{3*}$ & $l_6^3$ & $l_7^{4*}$ & $l_8^4$
      & $l_9^{5*}$ & $l_{10}^5$ & $l_{11}^5$ & $l_{12}^{6*}$
      & $l_{13}^6$ & $l_{14}^6$  \tabularnewline
     $\trailsave$
      &  &  &  & 
      &  &  &  &  
      &  &  &  &  
      &  & \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{1-UIP ($\lnot l_1$, $\lnot l_3$, $\lnot l_{12}$) and backtrack 2 ($\btL$)}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      &         &       &         &      
      &         &       &         &           
      &         &         \tabularnewline
      $\trailsave$
      &         &       &         &                     
      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
      & $l_9^*$ & $l_{10}$ & $l_{11}$ & 
      &         & 
      \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{Unit Prop $\lnot l_{12}$ ($\deepestLit$), $\trailsave$ not yet helpful}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $\smash{\lnot} l_{12}^2$  & $l_7^2$  & $l_9^2$    &      
      &         &       &         &           
      &         &         \tabularnewline
      $\trailsave$
      &         &       &         &                     
      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
      & $l_9^*$ & $l_{10}$ & $l_{11}$ &
      &         & 
      \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{Make decision}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $\smash{\lnot} l_{12}^2$  & $l_7^2$  & $l_9^2$ 
      & $l_5^{3*}$  &       &         &           
      &         &         \tabularnewline
      $\trailsave$
      &         &       &         &                     
      & $l_5^*$ & $l_6$ & $l_7^*$ & $l_8$
      & $l_9^*$ & $l_{10}$ & $l_{11}$ & 
      &         & 
      \tabularnewline
      \rowcolor{linen}
      \multicolumn{15}{l}{Now $\trailsave$ can be used to augment trail}\tabularnewline
      $\trail$ 
      & $l_1^{1*}$ & $l_2^1$ & $l_3^{2*}$ & $l_4^2$
      & $\lnot l_{12}^2$ \hfill & $l_7^2$  & $l_9^2$      
      & $l_5^{3*}$  & $l_6^{3}$  & $l_8^3$  & $l_{10}^3$          
      & $l_{11}^3$  &         \tabularnewline
      $\trailsave$
      &         &       &         &      
      &         &       &                              
      &         &       &         & 
      &         &       & 
    \end{tabular}
    \caption{Use of $\trailsave$ from
      Example~\ref{ex:trailsave}\label{fig:trailsave}. The literal's
      decision level is indicated in its superscript, and a $\mbox{}^*$
      superscript indicates that the literal is a decision.}
\end{figure}

\begin{example}
    \label{ex:trailsave} Figure~\ref{fig:trailsave} provides an
    example of how $\trailsave$ is used. Initially the literals $l_1$
    to $l_{14}$ are on the solver's $\trail$, and $\trailsave$ is
    empty. This is shown in the first two lines of the figure. In the
    figure the superscript on the literals indicates their decision
    level, and a superscripted $*$ indicates that the literal is a
    decision. Hence $l_1^{1*}$ indicates that $\dlevel{l_1}=1$ and
    that $l_1$ is a decision.

    Then a conflict is found at level $6$ and the 1-UIP clause
    $(\lnot l_1, \lnot l_3, \lnot l_{12})$ is learnt. Thus the solver
    will backtrack to level $2$, where it will add $\lnot l_{12}$ as a
    unit implicant. The next two lines show $\trail$ and $\trailsave$
    right after the backtrack to level $2$: the backtracked levels
    have been copied into $\trailsave$ omitting the conflict level $6$.

@@

As noted in the previous example unit propagation has to be rerun on
all saved literals added to $\trail$ from $\trailsave$. Thus our
technique, unlike \cbt, does not completely remove the overhead of
reproducing the trail on the solver's redescent. Nevertheless, trail
saving improves the efficiency of this redescent in three different
ways. First, by adding more forced literals to the trail before
continuing propagating the next literal, propagation can potentially
gain a quadratic speedup
\cite{DBLP:conf/sat/HickeyB19,DBLP:journals/jair/Gent13}. Second,
propagation does not need to examine the reason clause of the added
literals. If these literals were not added by \ust, propagation would
have to traverse each of these reason clauses to determine that they
have in fact become unit. Third, when a conflict is returned by
\ust all further propagations can be halted. The added literals and
their reasons will be sufficient to perform clause learning from the
conflict returned by \ust. Since trail saving can sometimes save
hundreds or thousands of literals at a time these savings can in sum
be significant.

@@

\subsection{Correctness}
Now we will prove that our use of $\trailsave$ preserves the \sat
solver's soundness. In particular, $\trailsave$ is only used in the
procedure \ust, in which it either adds new literals
to the solver's trail, or returns conflict clauses to the solver.
Hence, we only need to show that these new literals are in fact unit
implied and the conflicts are in fact falsified by the solver's trail.
\
Since both $\trail$ and $\trailsave$ are sequences of literals (with
associated reasons) we can consider their concatenation denoted as
$\trail + \trailsave$.
\begin{theorem}
    \label{thm:sound}
    If $\trail + \trailsave$ is \textbf{reason sound}
    (Section~\ref{sec:background}) then the following holds.  If the
    first $i$ literals on $\trailsave$ are all in $\trail$
    ($\forall j. 0\leq j< i. \trailsave[j]\in \trail$) and
    $\trailsave[i]=l$ is an implied literal with $\reasonsave{l} = C$,
    then $C$ has been made unit by $\trail$ implying $l$.
\end{theorem}
\noindent\emph{Proof:}
Since $\trail + \trailsave$ is reason sound, every literal in $C$
other than $l$ appears negated before $l$ in the sequence
$\trail + \trailsave$. Thus for $x\in C$ we have $\lnot x \in \trail$
or $\lnot x \in \trailsave[0]\ldots\trailsave[i-1]$. But in the later
case we also have $\lnot x \in \trail$. 
\whitebox

@@

\begin{theorem}
    \label{thm:rsound}
    $\trail+\trailsave$ is always reason sound during the operation of
    the solver.
\end{theorem}
\noindent \emph{Proof:} $\trailsave$ starts off being empty, so
$\trail+\trailsave = \trail$ is reason sound as it was generated by
the solver (4). In procedure \bt $\trail+\trailsave$ is
set to a trail that was previously generated by the solver (4). The solver
can add to $\trail$ by decisions and propagations without using
$\trailsave$. In this case $\trail' = \trail+\trail_{\mathit{new}}$,
and $\trail'$ is reason sound by (4), thus the new
$\trail'+\trailsave$ is reason sound by (3). Finally, in procedure
\ust either (a) literals at the front of
$\trailsave$ are discarded since they already appear on $\trail$, or
(b) literals are moved from $\trailsave$ to $\trail$. Under both of
these changes $\trail+\trailsave$ remains reason sound by (1) and (2).
\whitebox

@@

\paragraph{Saving the trail over multiple backtracks.}
It can often be the case that when the solver finds a conflict and
backtracks to $\btL$ it might immediately find a another conflict at
$\btL$ causing a further backtrack. In the procedure \bt every
backtrack causes $\trailsave$ to be overwritten. Hence, in these cases
most of the trail will not be saved---only the portion from the last
backtrack. Our first extension addresses this potential issue and also
provides more general trail saving in other contexts as well.

@@

\paragraph{Lookahead for Conflicts.}
In \ust we stop adding literals from $\trailsave$ to $\trail$ once we
reach a decision literal $d$ on $\trailsave$ that is not yet on
$\trail$ (line~\ref{ln:decnottrue} of \ust). This is done so that the
solver has full control over variable decisions without interference
from the trail saving mechanism (unlike the case with \cbt). However,
another option would be to force the solver to use $d$ as its next
decision literal, which would then allow us to further add all of
$d$'s implied literals on $\trailsave$ onto $\trail$. This can be done
for the first $k$ decisions on $\trailsave$ for any $k$. But in
general, we do not want to remove the solver's autonomy by forcing it
to make potentially different decisions than it might have wanted to.

@@

We experimented with this ``lookahead for conflicts'' idea using
various values of $k$. We found that $k=2$, i.e., forcing up to two
decisions from $\trailsave$ to be made by the solver if this yields a
conflict, often enhanced the solver's performance. Limiting the lookahead to
only one decision level of $\trailsave$ was not as good, and looking
ahead more than 2 decisions of $\trailsave$ also degraded
performance. This provides some evidence that taking too much
control away from the solver and forcing it to make too many decisions
from $\trailsave$ can lead to conflicts that are not as useful to the
solver.

\paragraph{Reason quality.}
The saved trail can be thought of as remembering the solver's recent
trajectory. Sometimes we want to follow the past trajectory, but
perhaps sometimes we do not. In particular, when adding literals from
$\trailsave{}$ to $\trail$ we can examine the quality of the saved
reasons to see if they are worth using. Once we encounter a literal
with a low quality saved reason we stop adding literals from
$\trailsave$ to the solver's trail. In particular, we can change
lines~\ref{ln:addToTrail} and \ref{ln:addToTrailNext} of \ust
to the following:\\
\begin{tabular}[t]{l}
23.5 \qquad \textbf{if} (lowQuality($\reasonsave{\litsave}$)) \textbf{break}\\
24.  \qquad $\trail.\mathrm{addToEnd(\litsave)}$\\
25.  \qquad $\reason{\litsave}\gets \reasonsave{\litsave}$
\end{tabular}

Note that the solver will still set the un-added literals as they are
unit implied by $\trail$, but it might be able to find better reasons
for these implicants. There is of course no guarantee that better
reasons will be found, but our empirical results show that sometimes
this does happen. We experimented with two quality metrics, clause
size and clause LBD, obtaining positive results with both. These
results also provides evidence against the argument given in
\cite{DBLP:journals/jsat/TakRH11} that changing literal reasons is not
impactful. With an appropriate clause quality metric the changing of literal
reasons can have an impact.

@@

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      maple           & 458          & 259 &  199  & 4746                             \\ \hline
      maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
      maple-trail-multipleBT-lookahead & \textbf{473} & \textbf{278} &  195  & \textbf{4584} \\ \hline
      maple-trail-multipleBT-lookahead-reason & 467 & 273 & 194 & 4668 \\ \hline
    \end{tabular}
    \caption{Table of results for MapleLCMDist.}
    \label{fig:maple}
\end{figure}
